Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-22 paper/ms.tex
final edits, added more refs

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        a_+\,\cos(d_+\,\tau) + b_+\,\sin(d_+\,\tau) +
        a_-\,\cos(d_-\,\tau) + b_-\,\sin(d_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to scale
inferences to large datasets for any model where the kernel function is a
general sum or product of \celerite\ terms.

@@

To demonstrate the application of \celerite, we will start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we will expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to measure
the power spectrum of a process when the data are generated from a \celerite\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is a interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

This model requires about a minute of computation time to find the maximum
likelihood parameters and then it takes about an hour to run the MCMC sampling
to convergence on a dual-CPU laptop.
This is more computationally intensive than traditional methods of measuring
asteroseismic oscillations but is is also orders of magnitude cheaper than the
same analysis using another GP solver.
An in depth discussion of the benefits of rigorous probabilistic inference of
asteroseismic parameters in the time domain is beyond the scope of this paper
but we hope to revisit this opportunity in the future.

@@

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/f$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \celerite\
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.

Although we have in mind applications of this fast method to model stellar
variability, the method is general for one-dimensional GP problems, and may be
applied to other problems.
Accreting black holes show time series which may be modeled by correlated
noise \citep{Kelly:2014};  indeed, this was the motivation for the original
technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic
variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
Outside of astronomy, this technique may have application to terrestrial
seismology \citep{Robinson:1967}.
The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
non-parametric data modeling \citep{Wilson:2013}.

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat aliases if the full posterior
is explored.  One drawback is that the number of $J$ terms to include is
unknown, but may be inferred with computation of the evidence, or an
approximation such as the Bayesian information criterion \citep{Kelly:2014}.
In general, the parameters of the \celerite\ terms are more interesting due
to their physical meaning rather than the overall shape of the power spectrum,
as shown in the foregoing examples.

>>>>

Finally, we note that the product of two terms of the form found inside the
sum in \eq{celerite-kernel} can also be re-written as a sum with updated
parameters
\begin{eqnarray}\eqlabel{product-rule}
k_j(\tau) \, k_k(\tau) =
    e^{-\tilde{c}\,\tau}\,[
        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
    ]
\end{eqnarray}
where
\begin{eqnarray}
    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
    \tilde{c} &=& c_j + c_k \\
    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
\end{eqnarray}
Therefore, the method described in the following section can be used to scale
inferences to large datasets for any model where the kernel function is a
general sum or product of \celerite\ terms.

@@

To demonstrate the application of \celerite, we will start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we will expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to measure
the power spectrum of a process when the data are generated from a \celerite\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is an interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

This model requires about a minute of computation time to find the maximum
likelihood parameters and then it takes about an hour to run the MCMC sampling
to convergence on a dual-CPU laptop.
This is more computationally intensive than traditional methods of measuring
asteroseismic oscillations but is is also orders of magnitude cheaper than the
same analysis using another GP solver.
An in-depth discussion of the benefits of rigorous probabilistic inference of
asteroseismic parameters in the time domain is beyond the scope of this paper
but we hope to revisit this opportunity in the future.

@@

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/f$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low and high frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \celerite\
will be important for transit detection \citep{ForemanMackey:2015}, transit timing 
\citep{Agol:2005,Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler 
beaming \citep{Loeb:2003,Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase 
functions \citep{Knutson:2007,Zucker:2007}, and more.

Although we have in mind applications of this fast method to model stellar
variability, the method is general for one-dimensional GP problems, and may be
applied to other problems.
Accreting black holes show time series which may be modeled by correlated
noise \citep{Kelly:2014};  indeed, this was the motivation for the original
technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic
variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
Outside of astronomy, this technique may have application to terrestrial
seismology \citep{Robinson:1967}.
%The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
%non-parametric data modeling \citep{Wilson:2013}.

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum \citep{Wilson:2013} avoids these difficulties:  it allows for the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat aliases if the full posterior
is explored.  One drawback is that the number of $J$ terms to include is
unknown, but may be inferred with computation of the evidence, or an
approximation such as the Bayesian information criterion \citep{Kelly:2014}.
In general, the parameters of the \celerite\ terms are more interesting due
to their physical meaning rather than the overall shape of the power spectrum,
as shown in the foregoing examples.
